Discussion Draft

### 2020-03-27 doc/Discussion.rmd
small clean up

@@

<Broad Scale crossover placement patterns, male telomeric, female centromeric,
review the typical landscape, one of the major broad scale patterns,
exceptions
consequences, (differences in variation, Sardell paper)>

@@

@@

While these (age, technical noise, ect) are potential sources of technical noise -- these results indecate this isn't due to technical noise or age effects.

@@

@@

1. the Chromosome axis length // chromatin organizatio in meiotisis

according to the tethered loop model of DSB formation -- the 'real estate' for COs are the DNA loops - the density of whch is conserved in the axis and loop formation (later the SC str), so the number of loops is equvilent to the length of the axis. When the total amount of DNA is kept constant as for a given Chromsome, there is a negative relationship between the loop length and axi length. **The chromosomes are converted to a 2D space (the axis length of a given chromosome is the limited factor for total CO number)**   This structure has been called 'lampbrush' by some (cite).

2. Only one of the models we examined predicts sexual dimorphism for axis length.

The modeifer models do not parameterize this since they do not models / are not built to consider single meiotic events.

**COM** predicts that axis length will be larger in oocytes due to the larger volumne -- this would help in the bouquet stage -- may be connected to the 'the search' during homolog pairing.

the **spindle based selection** model does not specific predictions about this aspect -- assumes that the 2D areas for chromosomes are proportaional across stages (during the decondensing stage between pachtyene and dipl? / metaphase) -- the chromosomes condenses uniformly across the length.

Another thing to point out / regaurding this model -- is the connection to sex difference in interference strength. Our results of conserved within strain raw IFD lengths - yet when the IFDs are controled for total SC length, we observed sexual dimorphism with males having longer IFD^normalized^ (stronger interference), mirror those observed in [@petkov2007].
The model / explaination proposed in that paper is that The chromatin structure which has longer loops (and shorter axis) will have more DNA bps folded / sequestered into the DNA loops. Thus the strength of interference will be stronger because with each step (count) along the 2D axis, more bases pairs are being skipped above the axis. These findings point out two things, This part of heterochiasmy generates testable prediction: First, the interference strength measured in chromsoome axis units (or SC) is generally conserved between sexes. Second the sex with longer axis and shorter loops will have weaker interference compared to the other sex. Or conversely, the sex with stronger linkage map based interference woud have chromatin organization with shorter axis compared to the other sex. There is a plefora of data of sexual dimorphism in linkage map based interference strength indeicating sexual dimorphism, but comparative cytological data is lacking (difference in perparness of some sexes and complications due to idenification of specific chromosomes between cells.)

@@

A thing about this study that we were very interested in was examining the faster evolution in males with the high recombining rates. BUT applying these above models to the traits connected to the faster male evolution is a bit tricker since none of these moedls were designed to explain within species variation in recombination rates. (most were designed to explain the general pattern of sexual dimorphism in meiotic recombination rates.)

@@

Our results demonstrate that total gwRR isn't the most conserved, (we observed 3 other traits more conserved sexual dimorphism examined above). Also our data shovariyng degrees of heterochiasmy magnitude (0 to gough). We the general pattern 2 strains attribute this pattern of reversed heterochiasmy to faster evolution of gwRR in two strains males. Which models predict dfferent rates of evolution between the sexes. Some of the models could explain a reversal of the direction of sexaul dimrohpsim, however most all of the predictions for faster evolution result in males having reduced recombination rates. Under the haploid selection model a  **reversal of the direction of heterochiasmy** would be accompanied by vast changes in haploid stages or degree of imprinting. Following the meiotic drive based model, females should be predicted to have larger linkage maps/gwRR since they are more exposed / weaker to true meiotic drivers. A reversal in heteroschiasmy is not clearly predicted by the SACE model (unless with the -- except under the conditions discribed below). Under the sindle selection model, the male (centrosome/effect SAC) metaphase could drive directional selection (and faster evolution) in males compared to female. However the mechanism/aspect of different fitness is not clear. Under the COM model, evolutionary predictions are not clear (explicit), but following the logic would come from changes in axis length (to generate another wave point) -- which might be due to meiocyte volumne.

# Evolution in Interference

@@

Our results indecated that interference strength has evolved with the genome wide recombiantion rates in the three strains with faster male evolution.

 This is unexpected because most literature and logical models are built / assume that weaker dense spacing of crossovers along chromosomes (ie. weaker interference) is how .

The history, models and empiracal examples are extensively reviewed in [@ottoPaysuer2019]. It is notable that borad discriptions / broad scale patterns for the correlation of gwRR and interference strength (gamma parameter) do fit the expected negative pattern [@ottoPaysuer2019].

(genome wide rate vs single bivalent measures)

**Caveats**
We considered carefully if this pattern was due to error and we are confident that this pattern is genuine.

There's no model that fits evolution of heterochiasmy and interference independantly. Under our undertsanding of the models, the relatioship of genome wide recombiantion rate and interference strength is assumed to be negative. We now review how some of the assumptions of typical modifer models might not fit realities of the meiotic pathway.

First, the **typical modifer models do not model single meiotic events**. (think about how pooling man meiotic events can skew patterns)
(crossover interference can be quantified from linkage maps (which are the average of many hundreds to thousands of meiotic events). It can also be discribed / measrued /quantified from single bivalets / chromosomes (either cytology of dense genotyping of single cells (cite). 
While it's widely agreed that they both reflect the same biological process, these two approaches are fundamentally different. (they are subject to distinct biases and noise).

Second, most modifer models use abstracted genomic space -- like an infite 2D line between markers. (space between markers is abstracted). This is useful for fitting the gamma model, but within the cell, the physical chromosomes's axis length is fitnite and crossovers cannot form beyond those bounds.

We conclude that under the haploid selection and meiotic driver model any prediction for evolution of interference strength follows from the 'logical' expectation of a negative correlation.

originally proposed/outlined by Trivers [] and refined /formalized in [@lenormand2003]. This model describes a system where gametes are competing or selection is acting in the haploid stage to cause difference in genome wide recombination rates between males and females.

 This is a modifier model of three loci, where the recombination rate between two loci is determined by a third locus (or modifier).

This model predicts the evolution of heterchiasmy -- through models (that build on selection to reduce recombiantion for the sex with geater varatince in fitness effect)

**SACE**
While the SACE model doesn't explecitly make predictions aof the out come for interference strength (or averge spacing of crossovers, )
We note that the main predicted pattern under this model of maintaining larger blocks of chromosomes in linkage can asrise from stronger interference / greater spacing of the distance between two foci (also maximize rbar).

 extends the lenomand focusing on two conditions. first that the stength of epistatsis in dependant on the cis or trans nature. the authors note that this would apply to gene archtechture (and increase with in cis ) -- where the regulatroy region and coding region of a gene evolve (together)

 second that the relative interference strengths differnt between male and female.  This can apply to (genes acting in sexual antagonism) or can be mediated via imprinted genes.

@@

**spindle based selection model**
There is a model that predicts this positive correlation between genome wide recombination rate and interference. The goldstein results (reviewed in).

@@

note about the differences in the patterns>  assumptions with gamma dstribution and linkage maps, (closer to infinte genetic space), less weight on crossover assurance - more weigth on 
(pooled SC/chromosome length gives strong correlation )

which is much different than when a single chromosome is examined (by itself), finite 2D length, crossover assurance, weaker correlation with ((the closer to chromosome level for SC legnth metric / summary, the weaker the correlation with crossover number))

2.  Review the assumptions for measuring CO interference in different ways.
The logical model: Crossover interference is a mechanism of suppressing crossovers, negative relationship between interference strength and gwRR. More crossovers can fit along the chromosomes and results in more crossovers overall.
Examples of empirical results supporting this pattern;  Otto Payseur, (Ruiz-Herrera? Bomblies?)
Quantifing using the gamma distribution: 
Caveats,  This doesn't make since for the general recombination landscape:   i) finite chromosome length and ii) limited range of crossovers per chromosomes (1 to 4).
Alternative interference characterization:  i) the number of crossovers are held constant and ii) transform / generalize the rec landscape, (focus on how the rec landscape would change instead of the genome wide rate) iii) assume crossover assurance.
	-The Goldstein model, models predicts a positive correlation with interference strength and higher recombination rate. A modifier which changes the probability of the recombination landscape having 0 crossovers to 2 crossovers, (driven by crossover assurance).

-Examples of empirical results which show this result / pattern
	-(cattle) comparison WITHIN sexes
	- Preliminary results for F2 cross (Hannah's work), (PWD vs CAST)
	(maybe more once I start looking closer)

This new model has **strengths**:
  - It matches fine scale differences in how DSB hotspots had different sex specific strengths with varied due to sex, epigenetc landscape and the gene architecture (Brick et al)

@@

<We can't distinguish if this is 3 independent instances of evolution in gwRR OR if the patterns are due to shared standing variation, incomplete lineage sorting, but these points are for describing the general patterns.>

>>>>

We focus first on the results for sexually dimorphic traits.

@@

@@

The conclusions from the models -- basically all of the models could account for differences in the general recombination landscape patterns.

@@

@@

None of the models make predictions for between cell variation -- so we draw conclusions from their preciations based on the strength of interference.

@@

**chromatin organization**

According to the tethered loop model of DSB formation (cite) -- the 'real estate' for COs are the DNA loops - the density of whch is conserved in the axis and loop formation (later the SC str), so the number of loops is equvilent to the length of the axis. **The chromosomes are converted to a 2D space (the axis length of a given chromosome is the limited factor for total CO number)**

2. Only one of the models we examined aligns with our observed results for sexual dimorphism for axis length.

The modifier models do not parameterize this since they do not models / are not built to consider single meiotic events.

**COM** predicts that axis length will be larger in oocytes due to the larger volume -- this would help in the bouquet stage -- may be connected to the 'the search' during homolog pairing.

the **spindle based selection** model does not specific predictions about this aspect -- assumes that the 2D areas for chromosomes are proportional across stages (during the decondensing stage between pachtyene and dipl? / metaphase) -- the chromosomes condenses uniformly across the length.

Another thing to point out / regarding this model -- is the connection to sex difference in interference strength. Our results of conserved within strain raw IFD lengths - yet when the IFDs are controlled for total SC length, we observed sexual dimorphism with males having longer IFD^normalized^ (stronger interference), mirror those observed in [@petkov2007].
The model / explanation proposed in that paper is that The chromatin structure which has longer loops (and shorter axis) will have more DNA bps folded / sequestered into the DNA loops. Thus the strength of interference will be stronger because with each step (count) along the 2D axis, more bases pairs are being skipped above the axis. These findings point out two things, This part of heterochiasmy generates testable prediction: First, the interference strength measured in chromosome axis units (or SC) is generally conserved between sexes. Second the sex with longer axis and shorter loops will have weaker interference compared to the other sex. Or conversely, the sex with stronger linkage map based interference would have chromatin organization with shorter axis compared to the other sex. There is a plethora of data of sexual dimorphism in linkage map based interference strength indicating sexual dimorphism, but comparative cytological data is lacking (difference in perparness of some sexes and complications due to identification of specific chromosomes between cells.)

@@

The faster evolution in males with the high recombining rates. BUT applying these above models to the traits connected to the faster male evolution is a bit trickier since none of these models were designed to explain within species variation in recombination rates. (most were designed to explain the general pattern of sexual dimorphism in meiotic recombination rates.)

@@

Our results demonstrate that gwRR is decoupled from other sexually dimorphic meotic traits, (we observed 3 other traits more conserved sexual dimorphism examined above). We the general pattern 2 strains attribute this pattern of reversed heterochiasmy to faster evolution of gwRR in two strains males.

Which models predict different rates of evolution between the sexes. Some of the models could explain a reversal of the direction of sexual dimrohpsim, however most all of the predictions for faster evolution result in males having reduced recombination rates. Under the haploid selection model a  **reversal of the direction of heterochiasmy** would be accompanied by vast changes in haploid stages or degree of imprinting. Following the meiotic drive based model, females should be predicted to have larger linkage maps/gwRR since they are more exposed / weaker to true meiotic drivers. A reversal in heteroschiasmy is not clearly predicted by the SACE model (unless with the -- except under the conditions described below). Under the spindle selection model, the male (centrosome/effect SAC) metaphase could drive directional selection (and faster evolution) in males compared to female. However the mechanism/aspect of different fitness is not clear. Under the COM model, evolutionary predictions are not clear (explicit), but following the logic would come from changes in axis length (to generate another wave point) -- which might be due to meiocyte volume.

# Evolution of Interference

@@

Our results indicated that interference strength has evolved with the genome wide recombination rates in the three strains with faster male evolution.

This is unexpected because most literature and logical models are built / assume that weaker dense spacing of crossovers along chromosomes (ie. weaker interference) is how. The history, models and empirical examples are extensively reviewed in [@ottoPaysuer2019]. Broad scale patterns for the correlation of gwRR and interference strength (gamma parameter) do fit the expected negative pattern [@ottoPaysuer2019].

There's no model that fits evolution of heterochiasmy and interference together. Under our undertsanding of the models, the relatioship of genome wide recombiantion rate and interference strength is assumed to be negative. We now review how some of the assumptions of typical modifer models might not fit realities of the meiotic pathway.

First, the **typical modifer models do not model single meiotic events**. We conclude that under the haploid selection and meiotic driver model any prediction for evolution of interference strength follows from the 'logical' expectation of a negative correlation.

This model predicts the evolution of heterchiasmy -through models (that build on selection to reduce recombiantion for the sex with geater varatince in fitness effect)

**SACE**
While the SACE model doesn't explecitly make predictions aof the out come for interference strength (or averge spacing of crossovers)
We note that the main predicted pattern under this model of maintaining larger blocks of chromosomes in linkage can asrise from stronger interference / greater spacing of the distance between two foci (also maximize rbar).

@@

@@

We propose that the spindle based selection model is the most parsimounious with our results and known recombination patterns from the literautre.

  - It matches fine scale differences in how DSB hotspots had different sex specific strengths with varied due to sex, epigenetc landscape and the gene architecture (Brick et al)

@@

<We can't distinguish if this is 3 independent instances of evolution in gwRR OR if the patterns are due to shared standing variation, incomplete lineage sorting, but these points are for describing the general patterns.>
<extends the lenomand focusing on two conditions. first that the stength of epistatsis in dependant on the cis or trans nature. the authors note that this would apply to gene archtechture (and increase with in cis ) -- where the regulatroy region and coding region of a gene evolve (together).  second that the relative interference strengths differnt between male and female.  This can apply to (genes acting in sexual antagonism) or can be mediated via imprinted genes. 
originally proposed/outlined by Trivers [] and refined /formalized in [@lenormand2003]. This model describes a system where gametes are competing or selection is acting in the haploid stage to cause difference in genome wide recombination rates between males and females. This is a modifier model of three loci, where the recombination rate between two loci is determined by a third locus (or modifier)>
<2.  Review the assumptions for measuring CO interference in different ways.
The logical model: Crossover interference is a mechanism of suppressing crossovers, negative relationship between interference strength and gwRR. More crossovers can fit along the chromosomes and results in more crossovers overall.
Examples of empirical results supporting this pattern;  Otto Payseur, (Ruiz-Herrera? Bomblies?)
Quantifing using the gamma distribution: 
Caveats,  This doesn't make since for the general recombination landscape:   i) finite chromosome length and ii) limited range of crossovers per chromosomes (1 to 4).
Alternative interference characterization:  i) the number of crossovers are held constant and ii) transform / generalize the rec landscape, (focus on how the rec landscape would change instead of the genome wide rate) iii) assume crossover assurance.
	-The Goldstein model, models predicts a positive correlation with interference strength and higher recombination rate. A modifier which changes the probability of the recombination landscape having 0 crossovers to 2 crossovers, (driven by crossover assurance)>
